Fast Bayesian Inference for Exoplanet Discovery in Radial Velocity Data

### 2014-12-31 Paper/ms.tex
More editing

The prior for the velocity semi-amplitudes $\{A_i\}$ is important and will affect the
inferred number of planets $N$. For example, if the prior allows for very
low amplitude planets to exist, the data will not rule this out, and the posterior
for $N$ will include the possibility of large numbers of planets. The prior
will influence how many of these low amplitude planets will exist: if we
believe there are many, and the data are uninformative about low amplitude
planets, then the posterior distribution for $N$ will also indicate that there
may be many low amplitude planets. However, their other properties, such as
their orbital periods, will not be well determined. The Beta prior for
eccentricity was suggested by Gregory (priv. comm) and is an approximation
to the inferred frequency distribution of eccentricities in the population
\citep{kipping}.

@@

We ran the MCMC on the simulated dataset to obtain samples from the
posterior distribution. We obtained 520 posterior samples.
The posterior distribution for $N$, the number of
planets, is shown in Figure~\ref{fig:fake_data_N}. The true number of planets,
7, is not the most probable value, but it does have substantial probability.
The posterior distribution suggests that $N$ could be anywhere from 6 to 10.
The posterior distribution for the periods $\{P_i\}_{i=1}^N$ is shown in
Figure~\ref{fig:fake_data_periods}. Because of the label switching degeneracy,
the posterior distribution for each period is identical, so we pooled the
samples for all periods. Defining the log-periods by
$S_i =  \log_{10}\left[P_i/(\textnormal{1 day})\right]$,
Figure~\ref{fig:fake_data_periods} is a Monte Carlo representation
of the mixture distribution
\begin{eqnarray}
f(S) &=& \sum_{N=0}^{10} p(N | D)\sum_{i=1}^N p(S_i | N, D).
\end{eqnarray}
The posterior distribution for the periods, shown in
Figure~\ref{fig:fake_data_periods}, shows that six of the true periods were
recovered, with probability close to 1. One period (with $\log_{10}P_i \approx 1.15$)
which was actually present was not ``detected'' because it had a very small
amplitude. We note that the posterior probability near this period should not be
precisely zero. There is also some
evidence for periods which did not actually exist, however the posterior
probabilities for these peaks are not close to 1.

@@

@@

\begin{figure}
\includegraphics[scale=0.45]{Figures/logl0.eps}
\caption{Log likelihood vs. enclosed prior mass (a standard output of Nested
Sampling algorithms) for the $\nu$ Oph analysis.
There are several phase transitions (concave-up regions) present. The one at
$\log(X) \approx -70$ separates models which contain additional signals from
models that do not. Without using Nested Sampling it would be more difficult
to mix between these two situations and calculate the posterior probability
for the existence of the additional signals.
\label{fig:logl0}}
\end{figure}

@@

The marginal likelihood for our model was
$\log(\mathcal{Z}) \approx -220.5$. Nested Sampling also allows for calculation
of the ``information'', or Kullback-Leibler divergence (a quantity in information
theory) from the prior to the posterior, which quantifies how much we learned about the parameters:
\begin{eqnarray}
\mathcal{H} &=& \int p(\theta | D) \log\left[\frac{p(\theta | D)}{p(\theta)}\right] \, d\theta
\end{eqnarray}
An intuitive interpretation of this quantity is the number of times the
prior distribution had to be compressed by a factor of $e$ (if the logarithm
in the formula is a natural logarithm) to get to the posterior distribution.
For the $\nu$ Oph data the information was $\mathcal{H} \approx 76.6$ nats
(natural units) or 111 bits, so the posterior is roughly $e^{-76.6}$ times the
volume of the prior.

@@

The posterior distribution for $N$ is shown in Figure~\ref{fig:gliese581_N},
and shows strong evidence for at least eight periods
($P(N \leq 7 | D) = 0.012$).
It is now recognised that
there are many possible sources for oscillations in a data set and not all
such oscillations should be claimed as planets \citep{2014Sci...345..440R}. 
Our model cannot distinguish between oscillations due to planets and
oscillations due to stellar activity: any oscillations found in the dataset
will be described as ``planets'' by the model. However it is interesting that
we find many more signals in the data than previous authors.

By inspecting the posterior distribution for the periods
(Figure~\ref{fig:gliese581_periods}), we see that only four of the periods
are well determined, corresponding to the known periods of Gliese 581 b, c, d,
and e. The other ``periods'' are very uncertain. It is possible that a
non-sinusoidal signal due to stellar activity is being modelled as several
periods \citep{astero}, and if the model were extended to include a ``stochastic''
oscillation \citep{gaussproc}, the number of periods detected may be reduced
substantially. An example model fit for Gliese 581 is shown in
Figure~\ref{fig:gliese581}.

>>>>

For the velocity semi-amplitudes $\{A_i\}$, we chose an exponential distribution
given the hyperparameter $\mu_A$. Our prior for $\mu_A$ spans many orders of
magnitude but expresses a slight preference for $\mu_A$ being of order unity,
using a Cauchy distribution. The prior for the semi-amplitudes
will influence how many of these low amplitude planets will be inferred: if we
believe there are many, and the data are uninformative about low amplitude
planets, then the posterior distribution for $N$ will also indicate that there
may be many low amplitude planets. However, their other properties, such as
their orbital periods, will not be well determined. The Beta prior for
eccentricity was suggested by Gregory (priv. comm) and is an approximation
to the inferred frequency distribution of eccentricities in the population
\citep{kipping}.

@@

We ran our algorithm on the simulated dataset to obtain samples from the
posterior distribution. We obtained 520 posterior samples.
The posterior distribution for $N$, the number of
planets, is shown in Figure~\ref{fig:fake_data_N}. The true number of planets,
7, is not the most probable value, but it does have substantial probability.
The posterior distribution suggests that $N$ could be anywhere from 6 to 10.

@@

The posterior distribution for the periods $\{P_i\}_{i=1}^N$ is shown in
Figure~\ref{fig:fake_data_periods}. Because of the label switching degeneracy,
the posterior distribution for each period is identical, so we pooled the
samples for all periods. Defining the log-periods by
$S_i =  \log_{10}\left[P_i/(\textnormal{1 day})\right]$,
Figure~\ref{fig:fake_data_periods} is a Monte Carlo representation
of the mixture distribution
\begin{eqnarray}
f(S) &=& \sum_{N=0}^{10} p(N | D)\sum_{i=1}^N p(S_i | N, D).
\end{eqnarray}
If a certain period is accurately measured (i.e. its probability of existence
is close to 1 and the posterior distribution for its period is very narrow)
then it will appear in Figure 3 with a height of $\sim 520$. If the uncertainty
in the period is larger than the histogram bin width then the peak will be
spread over several bins.

The posterior distribution for the periods, shown in
Figure~\ref{fig:fake_data_periods}, shows that six of the true periods were
recovered, with probability close to 1. One period (with $\log_{10}P_i \approx 1.15$)
which was actually present was not ``detected'' because it had a very small
amplitude. We note that the posterior probability near this period should not be
precisely zero. There is also some
evidence for periods which did not actually exist, however the posterior
probabilities for these peaks are not close to 1.

@@

\begin{figure}
\includegraphics[scale=0.45]{Figures/logl0.eps}
\caption{Log likelihood vs. enclosed prior mass (a standard output of Nested
Sampling algorithms) for the $\nu$ Oph analysis.
There are several phase transitions (concave-up regions) present. The small one at
$\log(X) \approx -70$ separates models which contain additional signals from
models that do not. Without using Nested Sampling it would be more difficult
to mix between these two situations and calculate the posterior probability
for the existence of the additional signals.
\label{fig:logl0}}
\end{figure}

@@

The marginal likelihood for our model was
$\log(\mathcal{Z}) \approx -220.5$. Nested Sampling also allows for calculation
of the ``information'', or Kullback-Leibler divergence (a quantity in information
theory) from the prior to the posterior, which quantifies how much we learned about the parameters:
\begin{eqnarray}
\mathcal{H} &=& \int p(\theta | D) \log\left[\frac{p(\theta | D)}{p(\theta)}\right] \, d\theta
\end{eqnarray}
An intuitive interpretation of this quantity is the number of times the
prior distribution had to be compressed by a factor of $e$ (if the logarithm
in the formula is a natural logarithm) to get to the posterior distribution.
For the $\nu$ Oph data the information was $\mathcal{H} \approx 76.6$ nats
(natural units) or 111 bits, so the posterior occupies about $e^{-76.6}$ times
the prior volume.

@@

The posterior distribution for $N$ is shown in Figure~\ref{fig:gliese581_N},
and shows strong evidence for at least eight periods
($P(N \leq 7 | D) = 0.012$). Some authors \citep[e.g.][]{2011A&A...528L...5T}
recommend that the probability of $N$ planets should be $\geq$ 150 times
greater than the probability of $N-1$ planets existing before making a
claim that $N$ planets have been definitively detected. Such a decision rule
is presumably equivalent to a utility function where false positives are much
worse than false negatives. We note that applying this rule to our results, we
would assert $N=6$, even though this has a tiny probability.

It is now recognised that
there are many possible sources for oscillations in a data set and not all
such oscillations should be claimed as planets \citep{2014Sci...345..440R}. 
Our model cannot distinguish between oscillations due to planets and
oscillations due to stellar activity: any oscillations found in the dataset
will be described as ``planets'' by the model. However it is interesting that
we find many more signals in the data than previous authors.
By inspecting the posterior distribution for the periods
(Figure~\ref{fig:gliese581_periods}), we see that only four of the periods
are well determined, corresponding to the known periods of Gliese 581 b, c, d,
and e. The other ``periods'' are very uncertain. It is quite likely that a
non-sinusoidal signal due to stellar activity is being modelled as several
periods \citep{astero}, and if the model were extended to include a ``stochastic''
oscillation \citep{gaussproc}, the number of periods detected may be reduced
substantially. An example model fit for Gliese 581 is shown in
Figure~\ref{fig:gliese581}.
